Transparency to hybrid open access invoicing through publisher-provided metadata: A longitudinal article-level study of Elsevier

### 2021-02-03 analysis/manuscript.Rmd
save tmp

Uptake and financial studies have OA policy implications for hybrid OA. Most prominently, the cOAlition S, an international network of research funders, released its widely discussed Plan S. In September 2018, the cOAlition S announced that from 2021 onwards, signatories would no longer fund articles in hybrid OA journals. This principle includes so-called "mirror" and sister journals--hybrid journals with fully OA counterparts or split into their fully OA and subscription parts—that have launched since Plan S was announced but makes an exception for hybrid journals with transformative agreements [@Frantsv_g_2019]. The Plan S refers to the Efficiency and Standards for Article Charges (ESAC) guidelines [@Geschuhn_2017] as critical component of transformative agreements, emphasizing the need for publisher-provided metadata to address the lack of transparency surrounding hybrid OA. However, studies confirm that publisher with such agreements lack essential data about hybrid OA uptake and invoicing, or are not able or willing to provide such information publicly and instantaneously [@Monaghan_2020; @Marques_2019], which, in turn, not just limit the monitoring of hybrid OA, but also research studies on its size and financial implications.

@@

First, we obtained hybrid journals through an Elsevier APC price list from May 2020 provided as captured by Matthias (2020). We only included hybrid journals that were published under the hybrid model throughout the entire period of our analysis. Hence, we excluded journals that flipped from OA to hybrid and from hybrid to OA. This approach allowed us to ensure that the article-level differences we observed were not due changes at the journal-level (i.e., increases or decreases in number). Overall, we identified 1,970 unique hybrid journals.

@@

Based on our findings, around a third of the articles were invoiced directly through OA agreements (e.g., research funders, national library consortia). Strikingly, sponsors based in the UK accounted for more than half of these , and a quarter was charged to sponsors from other European countries. This finding reiterates reports from previous studies that the UK's OA profile differs markedly from other countries (cf. ref, ref) and points to the effects of science policy. To increase OA to research publications, the UK implemented centralized APC funding and embraced hybrid OA as a transition model. Within five years of the publication of the Finch Report in 2012, the UK recorded an 18% increase in immediate OA, coupled with a rise in hybrid OA from 2.7% in 2012 to 15.4% of all articles in 2016 [@Jubb_2017], which can be attributed to the availability of APC funding from Research Councils UK (RCUK). Indeed, RCUK block grants have been the largest single source of OA funds in the UK—the Wellcome Trust, and more recently transformative agreements [@Tickel_2018; @Jubb_2017]. However, since then publishing expenditures of UK universities have been rising rapidly (ref), so that several universities stopped supporting hybrid OA in late 2018-2019 [@oxford_2019; @walker_2019; @Castro_2019; @birmingham]. This development might explain the slight decrease in hybrid articles invoiced to the UK our data showed around that time (see Figure \ref{fig:invoice_sponsor_country}).

>>>>

Uptake and financial studies have OA policy implications for hybrid OA. Most prominently, the cOAlition S, an international network of research funders, released its widely discussed Plan S. In September 2018, the cOAlition S announced that from 2021 onwards, signatories would no longer fund articles in hybrid OA journals. This principle includes so-called "mirror" and sister journals--hybrid journals with fully OA counterparts or split into their fully OA and subscription parts—that have launched since Plan S was announced but makes an exception for hybrid journals with transformative agreements [@Plan_s; @Frantsv_g_2019; @McNutt_2019; @Harrison_2019]. The Plan S refers to the Efficiency and Standards for Article Charges (ESAC) guidelines [@Geschuhn_2017] as critical component of transformative agreements, emphasizing the need for publisher-provided metadata to address the lack of transparency surrounding hybrid OA. However, studies confirm that publisher with such agreements lack essential data about hybrid OA uptake and invoicing, or are not able or willing to provide such information publicly and instantaneously [@Monaghan_2020; @Marques_2019], which, in turn, not just limit the monitoring of hybrid OA, but also research studies on its size and financial implications.

@@

First, we obtained hybrid journals through an Elsevier APC price list from May 2020 provided as captured by @Matthias_2019. We only included hybrid journals that were published under the hybrid model throughout the entire period of our analysis. Hence, we excluded journals that flipped from OA to hybrid and from hybrid to OA [@Matthias_2020]. This approach allowed us to ensure that the article-level differences we observed were not due changes at the journal-level (i.e., increases or decreases in number). Overall, we identified 1,970 unique hybrid journals.

@@

Based on our findings, around a third of the articles were invoiced directly through OA agreements (e.g., research funders, national library consortia). Strikingly, sponsors based in the UK accounted for more than half of these , and a quarter was charged to sponsors from other European countries. This finding reiterates reports from previous studies that the UK's OA profile differs markedly from other countries (cf. @Jubb_2015; @Jubb_2017) and points to the effects of science policy. To increase OA to research publications, the UK implemented centralized APC funding and embraced hybrid OA as a transition model. Within five years of the publication of the Finch Report in 2012, the UK recorded an 18% increase in immediate OA, coupled with a rise in hybrid OA from 2.7% in 2012 to 15.4% of all articles in 2016 [@Jubb_2017], which can be attributed to the availability of APC funding from Research Councils UK (RCUK). Indeed, RCUK block grants have been the largest single source of OA funds in the UK—the Wellcome Trust, and more recently transformative agreements [@Tickel_2018; @Jubb_2017]. However, since then publishing expenditures of UK universities have been rising rapidly (ref), so that several universities stopped supporting hybrid OA in late 2018-2019 [@oxford_2019; @walker_2019; @Castro_2019; @birmingham]. This development might explain the slight decrease in hybrid articles invoiced to the UK our data showed around that time (see Figure \ref{fig:invoice_sponsor_country}).

### 2021-02-03 analysis/manuscript.tex
save tmp

\title{Transparency to hybrid open access invoicing through publisher-provided metadata: A longitudinal article-level study of Elsevier}
\shorttitle{Hybrid OA Elsevier}
% \author{Najko Jahn}
% \affiliation{State and University Library Göttingen, University of Göttingen, Germany}
% \author{Lisa Matthias}
% \affiliation{tba}
% \author{Mikael Laakso}
% \affiliation{tba}
%

\abstract{The financial flows in scholarly journal publishing have become increasingly complex with the growth of open access (OA), and the lack of data and transparency into these flows has been persistent. Opaqueness has concerned especially hybrid OA, where subscription-based journals also publish OA content if an optional fee is paid. By leveraging Elsevier’s recently introduced article metadata concerning invoiced recipients, this study provides the first longitudinal publisher-level study of uptake and invoicing of hybrid OA articles. The results show that the share of OA articles in Elseviers hybrid journals has grown slowly, from 2.6\% of all articles on average in 2015 to 3.7\% in 2019. About a third of hybrid OA articles are invoiced directly through an agreement with research funders or other type of organisational sponsor (e.g. national library consortia), with the majority invoiced to authors. 83\% of agreement-invoiced hybrid OA articles are to European organisations, suggesting a strong international skewness. The study discovered a large discrepancy in choice of publication license between author-invoiced articles (dominantly CC BY-NC-ND) and agreement-invoiced articles (dominantly CC BY). The results demonstrate the value of publisher-provided metadata covering invoiced recipients of articles, which would be even more useful through wider adoption among publishers.}

\begin{document}
\maketitle

@@

Uptake and financial studies have OA policy implications for hybrid OA.
Most prominently, the cOAlition S, an international network of research
funders, released its widely discussed Plan S. In September 2018, the
cOAlition S announced that from 2021 onwards, signatories would no
longer fund articles in hybrid OA journals. This principle includes
so-called ``mirror'' and sister journals--hybrid journals with fully OA
counterparts or split into their fully OA and subscription parts---that
have launched since Plan S was announced but makes an exception for
hybrid journals with transformative agreements (Frantsvåg \& Strømme,
2019). The Plan S refers to the Efficiency and Standards for Article
Charges (ESAC) guidelines (Geschuhn \& Stone, 2017) as critical
component of transformative agreements, emphasizing the need for
publisher-provided metadata to address the lack of transparency
surrounding hybrid OA. However, studies confirm that publisher with such
agreements lack essential data about hybrid OA uptake and invoicing, or
are not able or willing to provide such information publicly and
instantaneously (Marques et al., 2019; Monaghan et al., 2020), which, in
turn, not just limit the monitoring of hybrid OA, but also research
studies on its size and financial implications.

@@

First, we obtained hybrid journals through an Elsevier APC price list
from May 2020 provided as captured by Matthias (2020). We only included
hybrid journals that were published under the hybrid model throughout
the entire period of our analysis. Hence, we excluded journals that
flipped from OA to hybrid and from hybrid to OA. This approach allowed
us to ensure that the article-level differences we observed were not due
changes at the journal-level (i.e., increases or decreases in number).
Overall, we identified 1,970 unique hybrid journals.

@@

Based on our findings, around a third of the articles were invoiced
directly through OA agreements (e.g., research funders, national library
consortia). Strikingly, sponsors based in the UK accounted for more than
half of these , and a quarter was charged to sponsors from other
European countries. This finding reiterates reports from previous
studies that the UK's OA profile differs markedly from other countries
(cf.~ref, ref) and points to the effects of science policy. To increase
OA to research publications, the UK implemented centralized APC funding
and embraced hybrid OA as a transition model. Within five years of the
publication of the Finch Report in 2012, the UK recorded an 18\%
increase in immediate OA, coupled with a rise in hybrid OA from 2.7\% in
2012 to 15.4\% of all articles in 2016 (Jubb et al., 2017), which can be
attributed to the availability of APC funding from Research Councils UK
(RCUK). Indeed, RCUK block grants have been the largest single source of
OA funds in the UK---the Wellcome Trust, and more recently
transformative agreements (Jubb et al., 2017; Tickel, 2018). However,
since then publishing expenditures of UK universities have been rising
rapidly (ref), so that several universities stopped supporting hybrid OA
in late 2018-2019 (Castro, 2019; University of Birmingham, n.d.;
University of Oxford, 2019; Walker, 2019). This development might
explain the slight decrease in hybrid articles invoiced to the UK our
data showed around that time (see Figure
\ref{fig:invoice_sponsor_country}).

@@

@@

@@

@@

>>>>

\title{Transparency to hybrid open access invoicing through publisher-provided metadata: A longitudinal article-level study of Elsevier}
\shorttitle{Hybrid OA Elsevier}
\author{Najko Jahn}
\affiliation{State and University Library Göttingen, University of Göttingen, Germany}
\author{Lisa Matthias}
\affiliation{}
\author{Mikael Laakso}

\abstract{The financial flows in scholarly journal publishing have become increasingly complex with the growth of open access (OA), and the lack of data and transparency into these flows has been persistent. Opaqueness has concerned especially hybrid OA, where subscription-based journals also publish OA content if an optional fee is paid. By leveraging Elsevier’s recently introduced article metadata concerning invoiced recipients, this study provides the first longitudinal publisher-level study of uptake and invoicing of hybrid OA articles. The results show that the share of OA articles in Elseviers hybrid journals has grown slowly, from 2.6\% of all articles on average in 2015 to 3.7\% in 2019. About a third of hybrid OA articles are invoiced directly through an agreement with research funders or other type of organisational sponsor (e.g. national library consortia), with the majority invoiced to authors. 83\% of agreement-invoiced hybrid OA articles are to European organisations, suggesting a strong international skewness. The study discovered a large discrepancy in choice of publication license between author-invoiced articles (dominantly CC BY-NC-ND) and agreement-invoiced articles (dominantly CC BY). The results demonstrate the value of publisher-provided metadata covering invoiced recipients of articles, which would be even more useful through wider adoption among publishers.
}

\begin{document}

\maketitle

@@

Uptake and financial studies have OA policy implications for hybrid OA.
Most prominently, the cOAlition S, an international network of research
funders, released its widely discussed Plan S. In September 2018, the
cOAlition S announced that from 2021 onwards, signatories would no
longer fund articles in hybrid OA journals. This principle includes
so-called ``mirror'' and sister journals--hybrid journals with fully OA
counterparts or split into their fully OA and subscription parts---that
have launched since Plan S was announced but makes an exception for
hybrid journals with transformative agreements (Frantsvåg \& Strømme,
2019; Harrison, 2019; McNutt, 2019; S, n.d.). The Plan S refers to the
Efficiency and Standards for Article Charges (ESAC) guidelines (Geschuhn
\& Stone, 2017) as critical component of transformative agreements,
emphasizing the need for publisher-provided metadata to address the lack
of transparency surrounding hybrid OA. However, studies confirm that
publisher with such agreements lack essential data about hybrid OA
uptake and invoicing, or are not able or willing to provide such
information publicly and instantaneously (Marques et al., 2019; Monaghan
et al., 2020), which, in turn, not just limit the monitoring of hybrid
OA, but also research studies on its size and financial implications.

@@

First, we obtained hybrid journals through an Elsevier APC price list
from May 2020 provided as captured by Matthias et al. (2019). We only
included hybrid journals that were published under the hybrid model
throughout the entire period of our analysis. Hence, we excluded
journals that flipped from OA to hybrid and from hybrid to OA (Matthias,
2020). This approach allowed us to ensure that the article-level
differences we observed were not due changes at the journal-level (i.e.,
increases or decreases in number). Overall, we identified 1,970 unique
hybrid journals.

@@

Based on our findings, around a third of the articles were invoiced
directly through OA agreements (e.g., research funders, national library
consortia). Strikingly, sponsors based in the UK accounted for more than
half of these , and a quarter was charged to sponsors from other
European countries. This finding reiterates reports from previous
studies that the UK's OA profile differs markedly from other countries
(cf.~Jubb et al. (2015); Jubb et al. (2017)) and points to the effects
of science policy. To increase OA to research publications, the UK
implemented centralized APC funding and embraced hybrid OA as a
transition model. Within five years of the publication of the Finch
Report in 2012, the UK recorded an 18\% increase in immediate OA,
coupled with a rise in hybrid OA from 2.7\% in 2012 to 15.4\% of all
articles in 2016 (Jubb et al., 2017), which can be attributed to the
availability of APC funding from Research Councils UK (RCUK). Indeed,
RCUK block grants have been the largest single source of OA funds in the
UK---the Wellcome Trust, and more recently transformative agreements
(Jubb et al., 2017; Tickel, 2018). However, since then publishing
expenditures of UK universities have been rising rapidly (ref), so that
several universities stopped supporting hybrid OA in late 2018-2019
(Castro, 2019; University of Birmingham, n.d.; University of Oxford,
2019; Walker, 2019). This development might explain the slight decrease
in hybrid articles invoiced to the UK our data showed around that time
(see Figure \ref{fig:invoice_sponsor_country}).

@@

\leavevmode\hypertarget{ref-Harrison_2019}{}%
Harrison, P. (2019). \emph{What are mirror journals, and can they offer
a new world of open access?} Elsevier B.V.
\url{https://web.archive.org/web/20210109033031/https://www.elsevier.com/connect/what-are-mirror-journals-and-can-they-offer-a-new-world-of-open-access}

@@

\leavevmode\hypertarget{ref-Jubb_2015}{}%
Jubb, M., Goldstein, S., Amin, M., Plume, A., Oeben, S., Aisati, M.,
Pinfield, S., Bath, P., Salter, J., Johnson, R., \& Fosci, M. (2015).
\emph{Monitoring the transition to open access: A report for the
universities uk open access co-ordination group}.
\url{https://web.archive.org/web/20190523005240/https://www.acu.ac.uk/research-information-network/monitoring-transition-to-open-access}

@@

\leavevmode\hypertarget{ref-Matthias_2020}{}%
Matthias, L. (2020). \emph{Publisher oa portfolios 2.0} (Version 2.0)
{[}Data set{]}. Zenodo. \url{https://doi.org/10.5281/zenodo.3841568}

\leavevmode\hypertarget{ref-Matthias_2019}{}%
Matthias, L., Jahn, N., \& Laakso, M. (2019). The two-way street of open
access journal publishing: Flip it and reverse it. \emph{Publications},
\emph{7}(2), 23. \url{https://doi.org/10.3390/publications7020023}

\leavevmode\hypertarget{ref-McNutt_2019}{}%
McNutt, M. (2019). Opinion: \\
textquotedblleftPlan s\\
textquotedblright falls short for society publishers\\
textemdashand for the researchers they serve. \emph{Proceedings of the
National Academy of Sciences}, \emph{116}(7), 2400--2403.
\url{https://doi.org/10.1073/pnas.1900359116}

@@

\leavevmode\hypertarget{ref-Plan_s}{}%
S. (n.d.). \emph{Plan s: Principles and implementation}.
\url{https://web.archive.org/web/20210129165306/https://www.coalition-s.org/addendum-to-the-coalition-s-guidance-on-the-implementation-of-plan-s/principles-and-implementation/}.
